---
title: "Archaeological Workflow with heuristR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Archaeological Workflow with heuristR}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[UTF-8]{inputenc}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

`heuristR` is designed to make Heurist scripting safer and easier, especially
for repeatable research tasks where you want programmatic access to records,
metadata, and controlled write operations.

This vignette uses a completely fictional archaeology database called
`demo_desert_survey`. The URLs, database name, user accounts, and record IDs
below are examples only.

## Example project

Imagine a regional archaeology project recording:

- sites
- projectile points
- ceramic sherd lots
- survey units
- places
- people and organizations

In Heurist, those concepts would usually be modeled as record types connected by
resource pointer fields. `heuristR` lets you inspect that structure, query
records, and perform safe read-modify-write updates from R.

## Connect to a database

Start by creating a session and logging in.

```{r eval = FALSE}
library(heuristR)

session <- heurist_session(
  base_url = "https://heurist.example.org/heurist",
  database = "demo_desert_survey"
)

session <- heurist_login(
  session,
  username = "field_director",
  password = Sys.getenv("HEURIST_PASSWORD")
)
```

For local development, it is usually best to keep credentials in a project
`.Renviron` file rather than in scripts.

## Inspect record types and fields

Before writing any automation, inspect the database structure.

```{r eval = FALSE}
rectypes <- heurist_rectypes(session)
fields <- heurist_fields(session)
structure <- heurist_structure(session)
```

For an archaeology database, you might expect record types such as `Site`,
`Projectile point`, `Survey Unit`, and `Place`, with fields for chronology,
material, geometry, site number, and linked records.

## Fetch records

You can retrieve a single record by ID:

```{r eval = FALSE}
site_42 <- heurist_get_record(session, 42)
```

You can also query a set of records using Heurist's query syntax:

```{r eval = FALSE}
late_sites <- heurist_find_records(
  session,
  q = 't:Site "Late Archaic" sortby:-m'
)
```

That is useful for reports such as:

- all sites updated this week
- all projectile points linked to a specific site
- all places within a survey zone

## Create a new record

Suppose the crew identifies a new locality named `North Ridge Wash`.

```{r eval = FALSE}
change_create <- heurist_create_record(
  session,
  rectype = "Place",
  details = list(
    "1" = list("0" = "North Ridge Wash")
  )
)

new_place_id <- change_create$record_id
```

`heurist_create_record()` returns a `heurist_change` object. That means the
operation can be rolled back later if this was a test or an accidental insert.

## Safely update an existing record

Heurist write endpoints can be destructive if you send incomplete payloads
directly. `heurist_patch_record()` avoids that by:

1. reading the current record
2. merging your requested change in R
3. sending a full replacement payload back to Heurist

Example:

```{r eval = FALSE}
change_patch <- heurist_patch_record(
  session,
  record_id = 42,
  details = list(
    "1" = list("0" = "North Ridge Site"),
    "1108" = list("0" = new_place_id)
  ),
  mode = "replace"
)
```

In an archaeological workflow, this is the safer way to add a linked `Place`,
correct a site title, or add a newly assigned field identifier while preserving
all unrelated data already stored on the record.

## Link related records

If you already know the target field and record IDs, `heurist_link_record()`
provides a clearer helper for resource pointers.

```{r eval = FALSE}
change_link <- heurist_link_record(
  session,
  record_id = 108,
  field_id = 1108,
  target_record_id = 42
)
```

That pattern fits common archaeological relationships such as:

- linking a projectile point to a site
- linking a survey unit to a place
- linking an artifact lot to a repository or collection

## Roll back a scripted change

Every high-level write helper returns enough information to undo the change.

```{r eval = FALSE}
heurist_rollback(change_patch)
heurist_rollback(change_link)
heurist_rollback(change_create)
```

Rollback is especially useful when you are:

- testing a migration script
- prototyping a batch cleanup
- validating field mappings on a staging database

## Low-level endpoints for advanced workflows

The package also exposes low-level wrappers when you need raw access:

```{r eval = FALSE}
raw_read <- heurist_raw_record_output(session, list(q = "t:Site"))
raw_edit <- heurist_raw_record_edit(session, list(a = "delete", ids = "9999"))
raw_meta <- heurist_raw_entity(session, list(a = "structure", entity = "all"))
```

These are useful for debugging or for building new helpers, but the
high-level functions are a better default for most research scripts.

## Suggested workflow for archaeology teams

For repeatable project work, a good pattern is:

1. define credentials in `.Renviron`
2. inspect structure with `heurist_rectypes()` and `heurist_fields()`
3. prototype reads with `heurist_find_records()`
4. perform updates with `heurist_patch_record()`
5. keep the returned `heurist_change` objects until the script is validated

That approach gives you a much safer path for maintaining site, artifact, and
place data in Heurist from R.
