---
title: "Spatial Data Workflows"
output:
  rmarkdown::html_vignette: default
  rmarkdown::pdf_document: default
vignette: >
  %\VignetteIndexEntry{Spatial Data Workflows}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[UTF-8]{inputenc}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

`heuristR` can return Heurist records as `sf` objects when the source records
contain geometry or coordinate data. This makes it possible to use standard R
spatial tools for filtering, plotting, export, and interactive map display.

The examples in this vignette are intentionally illustrative. They show the
intended workflow, but the exact record types, field names, and query strings
will vary across Heurist databases.

## When to use spatial output

Use `as_sf = TRUE` when you want to:

- map Heurist records in R
- filter records spatially with `sf`
- export Heurist geometry to GeoJSON or other GIS formats
- visualize records in `leaflet` or `mapview`

If a record includes spatial geometry, `heuristR` will try to convert that
geometry to an `sf` object. If `sf` is not installed and spatial data are
returned, the package will warn rather than silently discarding that fact.

## Connect to a Heurist database

```{r eval = FALSE}
library(heuristR)

session <- heurist_session(
  base_url = "https://heurist.example.org/heurist",
  database = "demo_desert_survey"
)

session <- heurist_login(
  session,
  username = "field_director",
  password = Sys.getenv("HEURIST_PASSWORD")
)
```

## Return records as sf

Suppose the database contains a `Site` record type with mapped locations. You
can request those records directly as an `sf` object:

```{r eval = FALSE}
sites_sf <- heurist_find_records(
  session,
  q = "t:Site",
  as_sf = TRUE
)
```

For a single record:

```{r eval = FALSE}
site_sf <- heurist_get_record(
  session,
  record_id = 42,
  as_sf = TRUE
)
```

The resulting object can then be used with standard `sf` functions:

```{r eval = FALSE}
class(sites_sf)
sf::st_geometry_type(sites_sf)
sf::st_crs(sites_sf)
```

## Inspect and clean geometry

Not every record in a real database will necessarily have valid or non-empty
geometry. A common first step is to isolate the features you can map.

```{r eval = FALSE}
sites_sf <- sites_sf[!sf::st_is_empty(sites_sf), ]
sites_sf <- sf::st_transform(sites_sf, 4326)
```

You can also look at the number of empty or valid geometries:

```{r eval = FALSE}
sum(sf::st_is_empty(sites_sf))
all(sf::st_is_valid(sites_sf))
```

## Make a leaflet map

If you want an interactive web map in R, `leaflet` is a natural fit:

```{r eval = FALSE}
library(leaflet)

leaflet(sites_sf) |>
  addProviderTiles(providers$CartoDB.Positron) |>
  addCircleMarkers(
    radius = 4,
    weight = 1,
    color = "#212529",
    fillColor = "#0d6efd",
    fillOpacity = 0.75,
    popup = ~rec_Title
  )
```

For polygon or mixed geometry data, `addPolygons()` or `addGeoJSON()` may be a
better fit than `addCircleMarkers()`.

## Make a quick map with mapview

For exploratory work, `mapview` can be even faster:

```{r eval = FALSE}
library(mapview)

mapview(sites_sf, zcol = "rec_Title")
```

This is a convenient option when you want a quick map to inspect what came back
from the database before building a more polished `leaflet` visualization.

## Filter spatial records before mapping

Once records are in `sf` format, you can use spatial operations directly.

Example: select only records within a bounding box:

```{r eval = FALSE}
bbox <- sf::st_as_sfc(
  sf::st_bbox(c(xmin = -111, ymin = 35, xmax = -109, ymax = 36), crs = 4326)
)

sites_in_bbox <- sites_sf[sf::st_intersects(sites_sf, bbox, sparse = FALSE), ]
```

Or clip to a study area polygon:

```{r eval = FALSE}
study_area <- sf::st_read("study-area.geojson")
sites_clipped <- sf::st_intersection(sites_sf, study_area)
```

## Export spatial data

Because the result is a normal `sf` object, export is straightforward:

```{r eval = FALSE}
sf::st_write(sites_sf, "sites.geojson", driver = "GeoJSON")
sf::st_write(sites_sf, "sites.gpkg", driver = "GPKG")
```

That makes it easy to move data from Heurist into GIS workflows in QGIS, ArcGIS,
or other R-based spatial analysis pipelines.

## Typical workflow

A common spatial workflow with `heuristR` looks like this:

1. log in with `heurist_session()` and `heurist_login()`
2. query a spatial record set with `heurist_find_records(..., as_sf = TRUE)`
3. remove empty geometries and transform CRS if needed
4. inspect or filter the result with `sf`
5. visualize it with `leaflet` or `mapview`
6. export to GeoJSON or another GIS-friendly format if needed

This pattern works well for archaeology, environmental history, cultural
heritage, and other spatially oriented research projects built in Heurist.
